Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-04-01 talk/icooolps2009/paper.tex
finish the review

When a hot loop in the user program is identified, tracing is started. The
tracing interpreter is invoked to start tracing the language interpreter that is
running the user program. Of course the tracing interpreter cannot actually
trace the execution of the C representation of the language interpreter. Instead
it takes the state of the execution of the language interpreter and starts
tracing using a bytecode representation of the language interpreter. That means
there are two "versions" of the language interpreter embedded in the final
executable of the VM: On the one hand it is there as executable machine code, on
the other hand as bytecode for the tracing interpreter. It also means that
tracing is costly as it incurs exactly a double interpretation overhead.

From then on things proceed like described in Section \ref{sect:tracing}. The
tracing interpreter tries to find a loop in the user program, if it found one it
will produce machine code for that loop and this machine code will be
immediately executed. The machine code is executed until a guard fails. Then the
execution should fall back to normal interpretation by the language interpreter.
This falling back is possibly a complex process, since the guard failure can
have occurred arbitrarily deep in a helper function of the language interpreter,
which would make it hard to rebuild the state of the language interpreter and
let it run from that point (e.g. this would involve building a potentially deep
C stack). Instead the falling back is achieved by a special \emph{fallback
interpreter} which runs the language interpreter and the user program from the
point of the guard failure. The fallback interpreter is essentially a variant of
the tracing interpreter that does not keep a trace. The fallback interpreter
runs until execution reaches a safe point where it is easy to let the C version
of the language interpreter resume its operation. Usually this means that the
fallback interpreter executes at most one bytecode operation of the language
interpreter. After the language interpreter takes over again, the whole process
starts again.

@@

\textbf{Trace Trees:} This paper ignored the problem of guards that fail in a
large percentage of cases because there are several equally likely paths through
a loop. Just falling back to interpretation in this case is not practicable.
Therefore we also start tracing from guards that failed many times and produce
machine code for that path, instead of always falling back to interpretation.

@@

@@

>>>>

When a hot user loop is identified, tracing is started. The
tracing interpreter is invoked to start tracing the language interpreter that is
running the user program. Of course the tracing interpreter cannot actually
trace the execution of the C representation of the language interpreter. Instead
it takes the state of the execution of the language interpreter and starts
tracing using a bytecode representation of the language interpreter. That means
there are two "versions" of the language interpreter embedded in the final
executable of the VM: on the one hand it is there as executable machine code, on
the other hand as bytecode for the tracing interpreter. It also means that
tracing is costly as it incurs exactly a double interpretation overhead.

From then on things proceed like described in Section \ref{sect:tracing}. The
tracing interpreter tries to find a loop in the user program, if it finds one it
will produce machine code for that loop and this machine code will be
immediately executed. The machine code is executed until a guard fails. Then the
execution should fall back to normal interpretation by the language interpreter.
This falling back is possibly a complex process, since the guard failure can
have occurred arbitrarily deep in a helper function of the language interpreter,
which would make it hard to rebuild the state of the language interpreter and
let it run from that point (e.g. this would involve building a potentially deep
C stack). Instead the falling back is achieved by a special \emph{fallback
interpreter} which runs the language interpreter and the user program from the
point of the guard failure. The fallback interpreter is essentially a variant of
the tracing interpreter that does not keep a trace. The fallback interpreter
runs until execution reaches a safe point where it is easy to let the C version
of the language interpreter resume its operation. Usually this means that the
fallback interpreter executes at most one bytecode operation of the language
interpreter. After the language interpreter takes over again, the whole process
starts again.

@@

\textbf{Trace Trees:} This paper ignored the problem of guards that fail in a
large percentage of cases because there are several equally likely paths through
a loop. Just falling back to interpretation in this case is not practicable.
\sout{
Therefore we also start tracing from guards that failed many times and produce
machine code for that path, instead of always falling back to interpretation. 
}
\anto{
Therefore, if we find a guard that fails often enough, we start tracing from
there and produce efficient machine code for that case, instead of alwayas
falling back to interpretation.
}

@@

\anto{XXX: should we say that virtualizables are very cool, that nobody else
  does that and that they are vital to get good performaces with python
  without sacrificing compatibility?}

@@

\anto{I would cite ourselves (maybe the JIT technical report?) and maybe
  psyco}
